Dissertation Proposal

### 2023-11-07 proposal.tex
pushed through

\begin{abstract}
  Misconceptions about core linguistic concepts like mutable
  variables, mutable compound data, and their interaction with scope
  and higher-order functions seem to be widespread. But how do we
  detect them, given that experts have blind spots and may not realize
  the myriad ways in which students can misunderstand programs?
  Furthermore, once identified, what can we do to correct them?

@@

\end{itemize}
This semantic core can be seen in languages from ``object-oriented''
languages like C\# and Java, to ``scripting'' languages like
JavaScript, Python, and Ruby, to ``functional'' languages like the ML
and Lisp families. Of course, there are sometimes restrictions (e.g.,
Java has restrictions on closures) and extensions (such as the
documented semantic oddities of JavaScript and Python
\citep{bernhardtWat2012, guhaEssenceJavascript2010,
politzPythonFullMonty2013, politzTestedSemanticsGetters2012}). Still,
this semantic core bridges many syntaxes, and understanding it helps
when transferring knowledge from old languages to new ones. In
recognition of this deep commonality, in this paper we choose to call
this the \emph{Standard Model of Languages} (SMoL).

Unfortunately, this combination of features appears to also be non-trivial for
programmers to understand. Consider the following scenario: to create a
calculator, we have to construct a callback that can be attached to
each button. The following Python program seems to achieve the
construction of these callbacks and the act of pushing the buttons in
order:
\begin{lstlisting}[language=Python]
button_list = []

@@

%   \item I implement these misconceptions as interpreters, which
%         find weaknesses in our manual analysis and hold potential for
%         generative use in the future.

@@

% There are some small differences.
% \cite{fleuryParameterPassingRules1991} identifies a dynamic scope misconception
% that is different from \miscon{FlatEnv}:
% \begin{description}
%   \item[\surmisedMiscon{CallerEnv}] Function values don't remember their environments.
%     When a function is called, the function body is evaluated in an
%     environment that extends from the \textit{caller's} environment.
% \end{description}
% I don't include \surmisedMiscon{CallerEnv} in our analysis because in our
% data, all wrong answers that can be explained by \surmisedMiscon{CallerEnv}
% are also explainable by \miscon{FlatEnv}.

% There are overlaps between our survey and
% theirs. For instance, our \surmisedMiscon{CallerEnv} is their No.~47.
% However, because their descriptions are brief, it is difficult to tell
% whether a misconception in their survey matches our misconceptions.
% Because
% they provide neither misinterpreters nor representative program-output
% pairs, it is difficult to determine the overlaps precisely (showing
% the value of providing these two machine-runnable descriptions).
% At any rate,
% we certainly find
% no equivalent of \miscon{FlatEnv}, \miscon{DeepClosure}, and
% \miscon{DefByRef} in their survey.

Known misconceptions are often detected with a kind of instruments
called \emph{concept inventories}
\citep{hestenesForceConceptInventory1992,taylorComputerScienceConcept2014}.
In terms of mechanics, a concept inventory is just an instrument
consisting of multiple-choice questions (MCQs), where each question
has one correct answer and several wrong ones. However, the wrong ones
are chosen with great care. Each one has been validated so that if a
student picks it, we can quite unambiguously determine \emph{what
misconception the student has}. For instance, if the question is
``What is \lstinline|sqrt(4)|?'', then \lstinline|37| is probably an
uninteresting wrong answer, but if people appear to confuse
square-roots with squares, then \lstinline|16| would be present as an
answer.\footnote{Concept inventories are thus useful in many settings.
For instance, an educator can use them with clickers to get quick
feedback from a class. If several students pick a specific wrong
answer, the educator not only knows they are wrong, but also has a
strong inkling of \emph{precisely what} misconception that group has
and can address it directly.}

@@

There is an extensive body of literature on tutoring systems
(\cite{vanlehnBehaviorTutoringSystems2006a} is a quality survey), and
indeed whole conferences are dedicated to them. I draw on this
literature. In particular, it is common in the literature to talk
about a ``two-loop'' architecture
\citep{vanlehnBehaviorTutoringSystems2006a} where the outer loop
iterates through ``tasks'' (i.e., educational activities) and the
inner loop iterates through UI events within a task. I follow the same
structure in my tutor (\Cref{s:tutor-uiux}).

@@

To answer this RQ, I have done some literature review on what
pedagogic techniques are helpful (\Cref{s:pedagogic-techniques}). My
current takeaway is that refutation text, case comparison, notional
machines, and language levels are helpful. I have created a tutor
(\Cref{s:system-tutor}) that incorporates all these ideas
(\Cref{s:tutor-uiux}) and have collected data from all populations. I
plan to explore more literature and to improve the tutor if I find new
helpful techniques.

@@

Each tutorial might be delivered as multiple sections to avoid forcing
students to continuesly work for more than 30 minutes.

@@

\paragraph{Panelty to Wrong Choices} To discourage students from
making a random choice, the tutor has been giving panelty. Designing
the strategy of assigning panelty is non-trivial:
\begin{itemize}
  \item A common strategy is to grade students by their correctness.
  However, grading by correctness is not suitable for a tutoring
  system because we do not mind students doing badly on earlier tasks
  as long as they eventually master the concepts. Actually, if
  students generally succeed in all tasks, the tutoring system is
  failing -- it does not teach anything to the students.
  \item A alternative common strategy is to grade by completion.
  However, grading by completion alone encourage students to rush
  through the tutorials and hence might encourage students to make
  random choices.
\end{itemize}
In short, I want to make students feel as follows:
\begin{itemize}
  \item They won't lose grade as long as they finish the tutorials.
  \item They should avoid giving wrong answers.
\end{itemize}
To encourage this feeling, I have been doing the following: when I
deployed the tutor in my institution, University 1, I graded by
completion\footnote{I encourage collaborators to grade by
completeness, but I have no real control over how they actually use
the tutor.}; when a student fails a task, the tutor gives an
\emph{extra} task:
\begin{enumerate}

@@

The current scope tutorial uses the following prompt
\begin{quote}
  When we see a variable reference, how do we find the corresponding
  declaration, if any?
\end{quote}
This prompt has a few problems:
\begin{itemize}
  \item It is unclear to adapt it to other tutorials.
  \item Students tend to give brief response possibly because the
  questions is asking for a lot and because students are tired when
  they reach the end of a tutorial.
\end{itemize}.

@@

If the \textbf{TaskSpecific} model best explains the data, I deduce
that some programs are more confusing and that we need further
investigation into the relationship between the tasks and
misconceptions. If the \textbf{PositionSpecific} model best explains
\emph{and} students perform better in later tasks, I deduce that the
tutor seeems to be correcting misconceptions. If studnets perform
worse in later taks, the tutor is potentially doing harm.

@@

The current tutor is not in the final shape yet. In particular, many
of the wrong answers correspond to more than one misconceptions; some
misconceptions are represented by only one question; the order of
interpreting tasks are not randomized. I plan to tackles the first two
differences by adding and removing programs. Adding randomization
requires me to carefully check the dependencies between current
explanations and goal sentences and resolve those dependencies if any.

@@

% Additionally, in this paper, we intentionally use the word
% mis\emph{conceptions} rather than mis\emph{take}. A mistake can happen
% for any reason (e.g., selecting the wrong answer from a menu). A
% misconception, however, implies a conceptual problem: the student has
% formed an incorrect concept in their head. For instance, they may
% think that mutable structures are copied on function calls, or that
% scope is resolved dynamically. This requires probing what they are
% thinking.

@@

%   \item we can get past the expert blind spot,

%   \item we have a sense of what misconceptions students have, and

%   \item we can generally associate wrong answers with specific
%         misconceptions, approaching a concept inventory.

@@

In Quizius, experts create a prompt; in our case, students were to
create small but ``interesting'' programs using the SMoL
language. Quizius shows this prompt to students and gathers their
answers. Each student is then shown a set of programs created by other
students and asked to predict (without running it) the value produced
by the program.\footnote{In the course (\Cref{s:populations}),
  students were given credit for using Quizius but not penalized for wrong
  answers, reducing their incentive to ``cheat'' by running
  programs. They were also told that doing so would diminish the value of
  their answers. Some students seemed to do so anyway, but most
  honored the directive.} Students are also asked to provide a rationale for why
they think it will produce that output.

@@

Prior work using Quizius \citep{saarinenHarnessingWisdomClasses2019}
finds that students do author programs that the experts did not
imagine. In our case, I seeded Quizius with programs from prior papers
(\Cref{s:background-misconceptions}), which gives the first few
students programs to respond to. However, I found that Quizius
significantly expanded the scope of our problems and misconceptions.
In our final instrument, most programs were directly or indirectly
inspired by the output of Quizius.

@@

\begin{table}[t]
  \centering
  \begin{tabularx}{\linewidth}{l X}
    \textbf{Question}                           &
    \textbf{Table of Answers}                       \\\hline
    \lstinputlisting{programs/circularity.txt}  & {
        \begin{tabularx}{\linewidth}{r X}
          $^\dag$50\% & \lstinline|#(#(2 #(2 3)) #(2 3))| \\
          *29\%       & \lstinline|x=#(x x)| or
          something similar. Both \lstinline|(left x)| and \lstinline|(right x)|
          are \lstinline|x| itself.                       \\
          $^\dag$16\% & \lstinline|Error|                 \\
          5\%         & Other                        \\
        \end{tabularx}
    }                                               \\\hline
    % \lstinputlisting{programs/mvec_in_mvec.txt} & {
    %     \begin{tabularx}{\linewidth}{r X}
    %       *62\%       & \lstinline|#(#(100 2 3 4) #(100 2 3 4))| \\
    %       $^\dag$26\% & \lstinline|#(#(1 2 3 4) #(100 2 3 4))|   \\
    %       6\%         & \texttt{Error}                      \\
    %       4\%         & \lstinline|#(#(1 2 3 4) #(1 2 3 4))|     \\
    %       2\%         & Other                               \\
    %     \end{tabularx}
    % }                                               \\\hline
    % \lstinputlisting{programs/vset_in_let.txt}  & {
    %     \begin{tabularx}{\linewidth}{r X}
    %       *59\%       & \lstinline|#(10)|  \\
    %       $^\dag$36\% & \lstinline|#(123)| \\
    %       5\%         & Other         \\
    %     \end{tabularx}
    % }                                               \\\hline
    \lstinputlisting{programs/var_as_arg.txt}   & {
        \begin{tabularx}{\linewidth}{r X}
          *65\%       & \texttt{12} \\
          $^\dag$31\% & \texttt{0}  \\
          4\%         & Other       \\
        \end{tabularx}
    }                                               \\\hline
    \lstinputlisting{programs/setset.txt}       & {
        \begin{tabularx}{\linewidth}{r X}
          *59\%       & Other: \texttt{5 7}   \\
          $^\dag$27\% & \texttt{6 7}          \\
          11\%        & \texttt{5 5}          \\
          2\%         & Other: \texttt{Error} \\
          1\%         & Other: \texttt{5 6}   \\
        \end{tabularx}
    }                                               \\\hline
  \end{tabularx}
  \caption{Some questions and student answers from \quizState. Each
    table row is a program and the (relative) frequencies of
    student answers. Answers that can be considered correct are marked
    with \texttt{*}. Wrong answers that we discuss are
    marked with $^\dag$.}
  \label{t:quizzes-data-state}
\end{table}
\Cref{t:quizzes-data-state} lists interesting programs after the
addition of state. These data suggest the students have aliasing-related misconceptions:
\begin{itemize}
  \item Up to 50\% of students think vectors are copied rather than
        aliased.
  \item 16\% of students think trying to construct and print a
        self-referring vector would error. (This program is ambiguous:
        \emph{constructing} works in most SMoL languages, but \emph{printing}
        may well cause a problem. These identified ambiguities are
        addressed in \Cref{s:tutor-refine-program-set}.)
  \item 31\% of students think a variable is aliased by a parameter if
        the two variables have the same name. Perhaps interestingly,
        fewer students (27\%) think the variable aliasing would
        happen if the two variables had different names.
\end{itemize}

@@

\begin{table}[t]
  \centering
  \begin{tabularx}{\linewidth}{l X}
    \textbf{Question}                                 &
    \textbf{Table of Answers}                             \\\hline
    \lstinputlisting{programs/fun_returns_lambda.txt} & {
        \begin{tabularx}{\linewidth}{r X}
          *82\%       & \lstinline|3|     \\
          $^\dag$17\% & \lstinline|Error| \\
          1\%         & Other        \\
        \end{tabularx}
    }                                                     \\\hline
    \lstinputlisting{programs/fun_and_state_1.txt}    & {
        \begin{tabularx}{\linewidth}{r X}
          *74\%       & \lstinline|4|            \\
          $^\dag$21\% & \lstinline|3|            \\
          6\%         & Other: \lstinline|Error| \\
        \end{tabularx}
    }                                                     \\\hline
    % \lstinputlisting{programs/fun_and_state_2.txt}    & {
    %     \begin{tabularx}{\linewidth}{r X}
    %       *88\%       & \lstinline|4|            \\
    %       $^\dag$10\% & \lstinline|3|            \\
    %       3\%         & Other: \lstinline|Error| \\
    %     \end{tabularx}
    % }                                                     \\\hline
    \lstinputlisting{programs/counter.txt}            & {
        \begin{tabularx}{\linewidth}{r X}
          *62\%       & \lstinline|1 1 2 3 2| \\
          $^\dag$34\% & \lstinline|1 1 1 1 1| \\
          4\%         & Other            \\
          1\%         & \lstinline|1 1 2 3 4| \\
        \end{tabularx}
    }                                                     \\\hline
  \end{tabularx}
  \caption{Some questions and student answers from \quizHof. Each
    table row is a program and the (relative) frequencies of
    student answers. Answers that can be considered correct are marked
    with \texttt{*}. Wrong answers that we discuss are
    marked with $^\dag$.}
  \label{t:quizzes-data-hof}
\end{table}
\Cref{t:quizzes-data-hof} lists interesting programs after the
addition of closures and higher-order functions. It extends what we
saw with loops in \Cref{s:intro}: that students have misconceptions
about their interaction with mutable variables:
\begin{itemize}
  \item 17\% of students think \texttt{lambda} functions can't refer
        to free variables.
  \item 21\% of the students think mutating a variable defined outside
        a lambda can't possibly change the behavior of the lambda.
  \item Student understanding of the
        \lstinline|let|-over-\lstinline|lambda| pattern is weak.
\end{itemize}

@@

  % fill gaps in goal sentences
  \item I felt the existing programs were insufficient to cover the
        ideas we wanted students to work through. For instance, SMoL
        Quizzes did not alias vectors using \lstinline|defvar|, so we
        added this program:

@@

  \item I deferred the local binding forms to the end, since we felt
        they were less essential. I therefore rewrote programs that use
        it to not depend on it.

        % \item I removed some questions that we deemed ``Lispy'' (e.g.,
        %   involving the difference between parenthetical and infix syntax).

@@

In addition to generating this set of programs, I also modified the
answer options. In addition to retaining the correct and incorrect
answers from the SMoL Quizzes (and constructing our best guess of
analogous incorrect answers for the new problems), I added more wrong
answers to for the reasons given in \Cref{s:interpreting-tasks}. In
general, I ended up increasing the number of choices substantially:
only 8 out of the 89 questions have three or four choices; 70
questions have 5--8 choices; and 11 questions have 9--14 choices.

@@

\section{The SMoL Tutor}
\label{s:system-tutor}

So far, we have focused on identifying problems. As noted earlier
(\Cref{s:smol-quizzes}), we provided students with refutation texts
after the quizzes, but it is unclear to what extent students read,
understood, or internalized these. I also wanted to determine whether
other populations run into these issues, and it was unclear whether we
would be able to administer passive quizzes to them.

Furthermore, while the quizzes may be a useful diagnostic, our goal is
not only to find faults but to improve the understanding of basic
programming language behavior. I believe (and presumably so does
anyone else who writes a formal semantics!)\ that an understanding of
these basic program behaviors is important, and even more so when it
comes to understanding concurrency, ownership, and other advanced
features. Even more simply, misunderstanding these clearly impacts
development and debugging time.

In response, we decided to create a \emph{tutor}: the SMoL Tutor. It is built
around our quiz instruments and the detected misconceptions. We
describe the Tutor from a user's perspective in \Cref{s:tutor-uiux}, and
explain how it was populated from the
SMoL Quizzes in \Cref{s:tutor-refine-program-set}.
I then discuss what
we learned from its data (\Cref{s:misconceptions-tutor}), and finally
evaluate its effectiveness as a \emph{tutor} (\Cref{s:effect-tutor}).

\subsection{The User Experience}
\label{s:tutor-uiux}

The Tutor covers five major topics, shown on the left in
\Cref{t:tutor-goal-sentences}. The larger topics are further broken
down into 2--3 modules. The goal was for students to spend at most
20--30 minutes per module. Our data show that in practice, students
spent about 9.8 (median) minutes.

Each tutorial consists of a sequence of
questions. Most questions in the Tutor are \newterm{interpreting
  questions}.\footnote{The Tutor also asked students to perform some
  other activities, like showing programs and asking for their heap
  content, which we do not cover in this paper.}
These questions (illustrated in
\Cref{f:smol-tutor-example}) are versions of the SMoL
Quizzes (obtained by the process described in \Cref{s:tutor-refine-program-set}).
In each of these questions, students are shown a program
and asked to predict the program's running result(s) by answering an
MCQ (with an ``Other'' option). After making a choice, students receive
feedback. If a student chooses incorrectly, they are (a) given an
explanation that is \emph{based on the misconception associated with
  that wrong answer} (or a generic one, if there is not a specific
misconception),
and then (b) asked to answer a
second question:
\begin{enumerate}

  \item The second question is semantically the same as the first, but
        with superficial changes (e.g., variable names, constants, and operators are
        changed) so that the student cannot immediately
        guess the answer.

  \item Instead of multiple-choice, students must \emph{type} the answer
        into a text box. (The Tutor normalizes text to accommodate variations.)
        This is
        intentional. First, we wanted to force reflection on the
        explanation just given, whereas with an
        MCQ, students could just guess. Second, we felt that students would
        find typing more onerous than clicking. In case students
        had just guessed on a question, our hope is that the penalty of having to
        type means, on \emph{subsequent} tasks, they
        would be more likely to pause and reflect before choosing.

@@

>>>>

\begin{abstract}
  Misconceptions about core linguistic concepts like mutable
  variables, mutable compound data, and their interaction with scope
  and higher-order functions seem to be widespread. But how do we
  detect them, given that experts have blind spots and may not realize
  the myriad ways in which students can misunderstand programs?
  Furthermore, once identified, what can I do to correct them?

@@

\end{itemize}
This semantic core can be seen in languages from ``object-oriented''
languages like C\# and Java, to ``scripting'' languages like
JavaScript, Python, and Ruby, to ``functional'' languages like the ML
and Lisp families. Of course, there are sometimes restrictions (e.g.,
Java has restrictions on closures) and extensions (such as the
documented semantic oddities of JavaScript and Python
\citep{bernhardtWat2012, guhaEssenceJavascript2010,
politzPythonFullMonty2013, politzTestedSemanticsGetters2012}). Still,
this semantic core bridges many syntaxes, and understanding it helps
when transferring knowledge from old languages to new ones. In
recognition of this deep commonality, in this paper I choose to call
this the \emph{Standard Model of Languages} (SMoL).

Unfortunately, this combination of features appears to also be non-trivial for
programmers to understand. Consider the following scenario: to create a
calculator, I have to construct a callback that can be attached to
each button. The following Python program seems to achieve the
construction of these callbacks and the act of pushing the buttons in
order:
\begin{lstlisting}[language=Python]
button_list = []

@@

%   \item I implement these misconceptions as interpreters, which
%         find weaknesses in my manual analysis and hold potential for
%         generative use in the future.

@@

% There are some small differences.
% \cite{fleuryParameterPassingRules1991} identifies a dynamic scope misconception
% that is different from \miscon{FlatEnv}:
% \begin{description}
%   \item[\surmisedMiscon{CallerEnv}] Function values don't remember their environments.
%     When a function is called, the function body is evaluated in an
%     environment that extends from the \textit{caller's} environment.
% \end{description}
% I don't include \surmisedMiscon{CallerEnv} in my analysis because in our
% data, all wrong answers that can be explained by \surmisedMiscon{CallerEnv}
% are also explainable by \miscon{FlatEnv}.

% There are overlaps between my survey and
% theirs. For instance, my \surmisedMiscon{CallerEnv} is their No.~47.
% However, because their descriptions are brief, it is difficult to tell
% whether a misconception in their survey matches my misconceptions.
% Because
% they provide neither misinterpreters nor representative program-output
% pairs, it is difficult to determine the overlaps precisely (showing
% the value of providing these two machine-runnable descriptions).
% At any rate,
% I certainly find
% no equivalent of \miscon{FlatEnv}, \miscon{DeepClosure}, and
% \miscon{DefByRef} in their survey.

Known misconceptions are often detected with a kind of instruments
called \emph{concept inventories}
\citep{hestenesForceConceptInventory1992,taylorComputerScienceConcept2014}.
In terms of mechanics, a concept inventory is just an instrument
consisting of multiple-choice questions (MCQs), where each question
has one correct answer and several wrong ones. However, the wrong ones
are chosen with great care. Each one has been validated so that if a
student picks it, I can quite unambiguously determine \emph{what
misconception the student has}. For instance, if the question is
``What is \lstinline|sqrt(4)|?'', then \lstinline|37| is probably an
uninteresting wrong answer, but if people appear to confuse
square-roots with squares, then \lstinline|16| would be present as an
answer.\footnote{Concept inventories are thus useful in many settings.
For instance, an educator can use them with clickers to get quick
feedback from a class. If several students pick a specific wrong
answer, the educator not only knows they are wrong, but also has a
strong inkling of \emph{precisely what} misconception that group has
and can address it directly.}

@@

There is an extensive body of literature on tutoring systems
(\cite{vanlehnBehaviorTutoringSystems2006a} is a quality survey), and
indeed whole conferences are dedicated to them. I draw on this
literature. In particular, it is common in the literature to talk
about a ``two-loop'' architecture
\citep{vanlehnBehaviorTutoringSystems2006a} where the outer loop
iterates through ``tasks'' (i.e., educational activities) and the
inner loop iterates through UI events within a task. I follow the same
structure in my tutor (\Cref{s:tutor}).

@@

To answer this RQ, I have done some literature review on what
pedagogic techniques are helpful (\Cref{s:pedagogic-techniques}). My
current takeaway is that refutation text, case comparison, notional
machines, and language levels are helpful. I have created a tutor that
incorporates all these ideas (\Cref{s:tutor}) and have collected
data from all populations. I plan to explore more literature and to
improve the tutor if I find new helpful techniques.

@@

Each tutorial might be delivered as multiple modules to avoid forcing
students to continuesly work for more than 30 minutes.

@@

\paragraph{Panelty to Wrong Choices} To discourage students from
making a random choice, the tutor has been giving panelty. Designing
the strategy of assigning panelty is non-trivial:
\begin{itemize}
  \item A common strategy is to grade students by their correctness.
  However, grading by correctness is not suitable for a tutoring
  system because I do not mind students doing badly on earlier tasks
  as long as they eventually master the concepts. Actually, if
  students generally succeed in all tasks, the tutoring system is
  failing -- it does not teach anything to the students.
  \item A alternative common strategy is to grade by completion.
  However, grading by completion alone encourage students to rush
  through the tutorials and hence might encourage students to make
  random choices.
\end{itemize}
In short, I want to make students feel as follows:
\begin{itemize}
  \item They won't lose grade as long as they finish the tutorials.
  \item They should avoid giving wrong answers.
\end{itemize}
To encourage this feeling, I have been doing the following: when I
deployed the tutor in my institution, University 1, I graded by
completion\footnote{I encourage collaborators to grade by
completeness, but I have no real control over how they actually use
the tutor.}; when a student fails a task, the tutor gives an
\emph{extra} task:
\begin{enumerate}

@@

The current scope tutorial uses the following prompt
\begin{quote}
  When I see a variable reference, how do I find the corresponding
  declaration, if any?
\end{quote}
This prompt has a few problems:
\begin{itemize}
  \item It is unclear to adapt it to other tutorials.
  \item Students tend to give brief response possibly because the
  questions is asking for a lot and because students are tired when
  they reach the end of a tutorial.
\end{itemize}.

@@

If the \textbf{TaskSpecific} model best explains the data, I deduce
that some programs are more confusing and that I need further
investigation into the relationship between the tasks and
misconceptions. If the \textbf{PositionSpecific} model best explains
\emph{and} students perform better in later tasks, I deduce that the
tutor seeems to be correcting misconceptions. If studnets perform
worse in later taks, the tutor is potentially doing harm.

@@

The current tutor is not in the final shape yet. In particular, many
of the wrong answers correspond to more than one misconceptions; some
misconceptions are represented by only one question; the order of
interpreting tasks are not randomized. I plan to tackles the first two
differences by adding and removing programs. Adding randomization
requires me to carefully check the dependencies between current
explanations and other texts in the tutor and resolve those
dependencies if any.

@@

% Additionally, in this paper, I intentionally use the word
% mis\emph{conceptions} rather than mis\emph{take}. A mistake can happen
% for any reason (e.g., selecting the wrong answer from a menu). A
% misconception, however, implies a conceptual problem: the student has
% formed an incorrect concept in their head. For instance, they may
% think that mutable structures are copied on function calls, or that
% scope is resolved dynamically. This requires probing what they are
% thinking.

@@

%   \item I can get past the expert blind spot,

%   \item I have a sense of what misconceptions students have, and

%   \item I can generally associate wrong answers with specific
%         misconceptions, approaching a concept inventory.

@@

In Quizius, experts create a prompt; in my case, students were to
create small but ``interesting'' programs using the SMoL
language. Quizius shows this prompt to students and gathers their
answers. Each student is then shown a set of programs created by other
students and asked to predict (without running it) the value produced
by the program.\footnote{In the course (\Cref{s:populations}),
  students were given credit for using Quizius but not penalized for wrong
  answers, reducing their incentive to ``cheat'' by running
  programs. They were also told that doing so would diminish the value of
  their answers. Some students seemed to do so anyway, but most
  honored the directive.} Students are also asked to provide a rationale for why
they think it will produce that output.

@@

Prior work using Quizius \citep{saarinenHarnessingWisdomClasses2019}
finds that students do author programs that the experts did not
imagine. In my case, I seeded Quizius with programs from prior papers
(\Cref{s:background-misconceptions}), which gives the first few
students programs to respond to. However, I found that Quizius
significantly expanded the scope of my problems and misconceptions.
In my final instrument, most programs were directly or indirectly
inspired by the output of Quizius.

@@

\begin{table}[t]
  \centering
  \begin{tabularx}{\linewidth}{l X}
    \textbf{Question}                           &
    \textbf{Table of Answers}                       \\\hline
    \lstinputlisting{programs/circularity.txt}  & {
        \begin{tabularx}{\linewidth}{r X}
          $^\dag$50\% & \lstinline|#(#(2 #(2 3)) #(2 3))| \\
          *29\%       & \lstinline|x=#(x x)| or
          something similar. Both \lstinline|(left x)| and \lstinline|(right x)|
          are \lstinline|x| itself.                       \\
          $^\dag$16\% & \lstinline|Error|                 \\
          5\%         & Other                        \\
        \end{tabularx}
    }                                               \\\hline
    % \lstinputlisting{programs/mvec_in_mvec.txt} & {
    %     \begin{tabularx}{\linewidth}{r X}
    %       *62\%       & \lstinline|#(#(100 2 3 4) #(100 2 3 4))| \\
    %       $^\dag$26\% & \lstinline|#(#(1 2 3 4) #(100 2 3 4))|   \\
    %       6\%         & \texttt{Error}                      \\
    %       4\%         & \lstinline|#(#(1 2 3 4) #(1 2 3 4))|     \\
    %       2\%         & Other                               \\
    %     \end{tabularx}
    % }                                               \\\hline
    % \lstinputlisting{programs/vset_in_let.txt}  & {
    %     \begin{tabularx}{\linewidth}{r X}
    %       *59\%       & \lstinline|#(10)|  \\
    %       $^\dag$36\% & \lstinline|#(123)| \\
    %       5\%         & Other         \\
    %     \end{tabularx}
    % }                                               \\\hline
    \lstinputlisting{programs/var_as_arg.txt}   & {
        \begin{tabularx}{\linewidth}{r X}
          *65\%       & \texttt{12} \\
          $^\dag$31\% & \texttt{0}  \\
          4\%         & Other       \\
        \end{tabularx}
    }                                               \\\hline
    \lstinputlisting{programs/setset.txt}       & {
        \begin{tabularx}{\linewidth}{r X}
          *59\%       & Other: \texttt{5 7}   \\
          $^\dag$27\% & \texttt{6 7}          \\
          11\%        & \texttt{5 5}          \\
          2\%         & Other: \texttt{Error} \\
          1\%         & Other: \texttt{5 6}   \\
        \end{tabularx}
    }                                               \\\hline
  \end{tabularx}
  \caption{Some questions and student answers from \quizState. Each
    table row is a program and the (relative) frequencies of
    student answers. Answers that can be considered correct are marked
    with \texttt{*}. Wrong answers that I discuss are
    marked with $^\dag$.}
  \label{t:quizzes-data-state}
\end{table}
\Cref{t:quizzes-data-state} lists interesting programs after the
addition of state. These data suggest the students have aliasing-related misconceptions:
\begin{itemize}
  \item Up to 50\% of students think vectors are copied rather than
        aliased.
  \item 16\% of students think trying to construct and print a
        self-referring vector would error. (This program is ambiguous:
        \emph{constructing} works in most SMoL languages, but \emph{printing}
        may well cause a problem. These identified ambiguities are
        addressed in \Cref{s:tutor-refine-program-set}.)
  \item 31\% of students think a variable is aliased by a parameter if
        the two variables have the same name. Perhaps interestingly,
        fewer students (27\%) think the variable aliasing would
        happen if the two variables had different names.
\end{itemize}

@@

\begin{table}[t]
  \centering
  \begin{tabularx}{\linewidth}{l X}
    \textbf{Question}                                 &
    \textbf{Table of Answers}                             \\\hline
    \lstinputlisting{programs/fun_returns_lambda.txt} & {
        \begin{tabularx}{\linewidth}{r X}
          *82\%       & \lstinline|3|     \\
          $^\dag$17\% & \lstinline|Error| \\
          1\%         & Other        \\
        \end{tabularx}
    }                                                     \\\hline
    \lstinputlisting{programs/fun_and_state_1.txt}    & {
        \begin{tabularx}{\linewidth}{r X}
          *74\%       & \lstinline|4|            \\
          $^\dag$21\% & \lstinline|3|            \\
          6\%         & Other: \lstinline|Error| \\
        \end{tabularx}
    }                                                     \\\hline
    % \lstinputlisting{programs/fun_and_state_2.txt}    & {
    %     \begin{tabularx}{\linewidth}{r X}
    %       *88\%       & \lstinline|4|            \\
    %       $^\dag$10\% & \lstinline|3|            \\
    %       3\%         & Other: \lstinline|Error| \\
    %     \end{tabularx}
    % }                                                     \\\hline
    \lstinputlisting{programs/counter.txt}            & {
        \begin{tabularx}{\linewidth}{r X}
          *62\%       & \lstinline|1 1 2 3 2| \\
          $^\dag$34\% & \lstinline|1 1 1 1 1| \\
          4\%         & Other            \\
          1\%         & \lstinline|1 1 2 3 4| \\
        \end{tabularx}
    }                                                     \\\hline
  \end{tabularx}
  \caption{Some questions and student answers from \quizHof. Each
    table row is a program and the (relative) frequencies of
    student answers. Answers that can be considered correct are marked
    with \texttt{*}. Wrong answers that I discuss are
    marked with $^\dag$.}
  \label{t:quizzes-data-hof}
\end{table}
\Cref{t:quizzes-data-hof} lists interesting programs after the
addition of closures and higher-order functions. It extends what we
saw with loops in \Cref{s:intro}: that students have misconceptions
about their interaction with mutable variables:
\begin{itemize}
  \item 17\% of students think \texttt{lambda} functions can't refer
        to free variables.
  \item 21\% of the students think mutating a variable defined outside
        a lambda can't possibly change the behavior of the lambda.
  \item Student understanding of the
        \lstinline|let|-over-\lstinline|lambda| pattern is weak.
\end{itemize}

@@

  % fill gaps in goal sentences
  \item I felt the existing programs were insufficient to cover the
        ideas I wanted students to work through. For instance, SMoL
        Quizzes did not alias vectors using \lstinline|defvar|, so we
        added this program:

@@

  \item I deferred the local binding forms to the end, since I felt
        they were less essential. I therefore rewrote programs that use
        it to not depend on it.

        % \item I removed some questions that I deemed ``Lispy'' (e.g.,
        %   involving the difference between parenthetical and infix syntax).

@@

In addition to generating this set of programs, I also modified the
answer options. In addition to retaining the correct and incorrect
answers from the SMoL Quizzes (and constructing my best guess of
analogous incorrect answers for the new problems), I added more wrong
answers to for the reasons given in \Cref{s:interpreting-tasks}. In
general, I ended up increasing the number of choices substantially:
only 8 out of the 89 questions have three or four choices; 70
questions have 5--8 choices; and 11 questions have 9--14 choices.

@@

\section{The SMoL Tutor}%
\label{s:tutor}

% So far, I have focused on identifying problems. As noted earlier
% (\Cref{s:smol-quizzes}), I provided students with refutation texts
% after the quizzes, but it is unclear to what extent students read,
% understood, or internalized these. I also wanted to determine whether
% other populations run into these issues, and it was unclear whether we
% would be able to administer passive quizzes to them.

% Furthermore, while the quizzes may be a useful diagnostic, my goal is
% not only to find faults but to improve the understanding of basic
% programming language behavior. I believe (and presumably so does
% anyone else who writes a formal semantics!)\ that an understanding of
% these basic program behaviors is important, and even more so when it
% comes to understanding concurrency, ownership, and other advanced
% features. Even more simply, misunderstanding these clearly impacts
% development and debugging time.

% In response, I decided to create a \emph{tutor}: the SMoL Tutor. It is built
% around my quiz instruments and the detected misconceptions. We
% describe the Tutor from a user's perspective in \Cref{s:tutor}, and
% explain how it was populated from the
% SMoL Quizzes in \Cref{s:tutor-refine-program-set}.
% I then discuss what
% I learned from its data (\Cref{s:misconceptions-tutor}), and finally
% evaluate its effectiveness as a \emph{tutor} (\Cref{s:effect-tutor}).

% \subsection{The User Experience}

This section describe the current status of the tutor.

The Tutor includes five tutorials, shown on the left in
\Cref{t:tutor-goal-sentences}. The larger topics are further broken
down into 2--3 modules. The goal was for students to spend at most
20--30 minutes per module. my data show that in practice, students
spent about 9.8 (median) minutes.

\Cref{f:smol-tutor-example} illustrates an interpreting task as
descrbed in \Cref{s:interpreting-tasks}. The Tutor also asked students
to perform some other activities, like showing programs and asking for
their heap content, which I do not cover in this paper.

% Each tutorial consists of a sequence of questions. Most questions in
% the Tutor are \newterm{interpreting questions}.\footnote{The Tutor
% also asked students to perform some other activities, like showing
% programs and asking for their heap content, which I do not cover in
% this paper.} These questions (illustrated in
% \Cref{f:smol-tutor-example}) are versions of the SMoL Quizzes
% (obtained by the process described in
% \Cref{s:tutor-refine-program-set}). In each of these questions,
% students are shown a program and asked to predict the program's
% running result(s) by answering an MCQ (with an ``Other'' option).
% After making a choice, students receive feedback. If a student chooses
% incorrectly, they are (a) given an explanation that is \emph{based on
% the misconception associated with that wrong answer} (or a generic
% one, if there is not a specific misconception), and then (b) asked to
% answer a second question:
% \begin{enumerate}

%   \item The second question is semantically the same as the first, but
%         with superficial changes (e.g., variable names, constants, and operators are
%         changed) so that the student cannot immediately
%         guess the answer.

%   \item Instead of multiple-choice, students must \emph{type} the answer
%         into a text box. (The Tutor normalizes text to accommodate variations.)
%         This is
%         intentional. First, I wanted to force reflection on the
%         explanation just given, whereas with an
%         MCQ, students could just guess. Second, I felt that students would
%         find typing more onerous than clicking. In case students
%         had just guessed on a question, my hope is that the penalty of having to
%         type means, on \emph{subsequent} tasks, they
%         would be more likely to pause and reflect before choosing.

@@

% \paragraph{Confirmed Misconceptions}
